Predicting Thyroid Cancer

Libraries
```{r}
library(ggplot2) # for all the graphs/ plots
library(tidyverse)
```

Load the dataset
```{r}
data <- read.csv("breast-cancer.csv", stringsAsFactors = TRUE)
# summary(data)
# str(data)
# colnames(data)
```

Categorical variables as factors
```{r}
categorical_vars <- c("diagnosis")
data <- data %>%
  mutate(across(all_of(categorical_vars), as.factor))
```

Continuous variables as numeric
```{r}
numeric_vars <- c("radius_mean", "texture_mean", "perimeter_mean", "area_mean", 
                  "smoothness_mean", "compactness_mean", "concavity_mean", 
                  "concave.points_mean", "symmetry_mean", "fractal_dimension_mean",
                  "radius_se", "texture_se", "perimeter_se", "area_se",
                  "smoothness_se", "compactness_se", "concavity_se",
                  "concave.points_se", "symmetry_se", "fractal_dimension_se",
                  "radius_worst", "texture_worst", "perimeter_worst",
                  "area_worst", "smoothness_worst", "compactness_worst",
                  "concavity_worst", "concave.points_worst", "symmetry_worst",
                  "fractal_dimension_worst")
data <- data %>%
  mutate(across(all_of(numeric_vars), as.numeric))
```

Check for missing values, there are no missing values.
```{r}
colSums(is.na(data))
```

Create boxplots for all the variables
```{r}
# Distribution of Diagnosis
ggplot(data, aes(x = diagnosis, fill = diagnosis)) + 
  geom_bar() +
  theme_minimal() +
  ggtitle("Diagnosis Distribution")

# Plot for each numeric variable
for (var in numeric_vars) {
  
  # Boxplot comparing benign vs malignant
  boxplots <- ggplot(data, aes_string(x = "diagnosis", y = var, fill = "diagnosis")) +
    geom_boxplot() +
    theme_minimal() +
    ggtitle(paste(var, "by Diagnosis"))
  print(boxplots)
}
```



Set Benign and Malignant outcomes as 0,1.
There are 357 benign cases and 212 malignant cases in the data set.
```{r}
levels(data$diagnosis)[levels(data$diagnosis) %in% c("Benign")] <- "0"
levels(data$diagnosis)[levels(data$diagnosis) %in% c("Malignant")] <- "1"
summary(data$diagnosis)
```




Check correlation between numerical predictors
```{r}
# Compute the correlation matrix for numeric variables
cor_matrix <- cor(data[, numeric_vars])
print(cor_matrix)


```



```{r}
# Standardise numeric variables using scale() in base R
# data[ , numeric_vars] <- scale(data[ , numeric_vars])
# library(dplyr)
# 
# data <- data %>%
#   mutate(across(all_of(numeric_vars), ~ as.numeric(scale(.))))
```



```{r}
# GLM on full data
glm_model <- glm(diagnosis ~ . - id, data = data, family = binomial)

# Display model summary
summary(glm_model)
```





