Predicting Thyroid Cancer

Load the dataset
```{r}
thyroid_data <- read.csv("thyroid_cancer_risk_data.csv", stringsAsFactors = TRUE)
summary(thyroid_data)
```

Categorical variables as factors
```{r}
categorical_var <- c("Gender", "Country", "Ethnicity", "Family_History", "Radiation_Exposure", "Iodine_Deficiency", "Smoking", "Obesity", "Diabetes", "Thyroid_Cancer_Risk", "Diagnosis")
thyroid_data <- thyroid_data %>%
  mutate(across(all_of(categorical_var), as.factor))
```

Continuous variables as numeric
```{r}
numerical_var <- c("Age", "TSH_Level", "T3_Level", "T4_Level", "Nodule_Size")
thyroid_data <- thyroid_data %>%
  mutate(across(all_of(numerical_var), as.numeric))
```

Check for missing values, there are no missing values.
```{r}
colSums(is.na(thyroid_data))
```

Set Benign and Malignant outcomes as 0,1.
There are 163196 benign cases and 49495 malignant cases in the data set.
```{r}
# levels(thyroid_data$Diagnosis)[levels(thyroid_data$Diagnosis) %in% c("Benign")] <- "0"
# levels(thyroid_data$Diagnosis)[levels(thyroid_data$Diagnosis) %in% c("Malignant")] <- "1"
# summary(thyroid_data$Diagnosis)
```


Visually explore key themes using boxplots for the categorical predictors
```{r}
#library(ggplot2)

# Distribution of diagnosis
ggplot(thyroid_data, aes(x = Diagnosis, fill = Diagnosis)) + 
  geom_bar() + 
  theme_minimal() + 
  ggtitle("Diagnosis Distribution")

ggplot(thyroid_data, aes(x = Gender, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Gender and Diagnosis")
  
ggplot(thyroid_data, aes(x = Country, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Country and Diagnosis")
  
ggplot(thyroid_data, aes(x = Ethnicity, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Ethnicity and Diagnosis")
  
ggplot(thyroid_data, aes(x = Family_History, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Family History and Diagnosis")

ggplot(thyroid_data, aes(x = Radiation_Exposure, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Radiation Exposure and Diagnosis")
  
ggplot(thyroid_data, aes(x = Iodine_Deficiency, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Iodine Deficiency and Diagnosis")
  
ggplot(thyroid_data, aes(x = Smoking, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Smoking and Diagnosis")

ggplot(thyroid_data, aes(x = Obesity, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Obesity and Diagnosis")

ggplot(thyroid_data, aes(x = Diabetes, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Diabetes and Diagnosis")

ggplot(thyroid_data, aes(x = Thyroid_Cancer_Risk, fill = Diagnosis)) + 
  geom_bar(position = "fill") + 
  ggtitle("Thyroid_Cancer_Risk and Diagnosis")

```


Boxplots for the numerical predictors
```{r}
# Boxplot for TSH, T3, and T4 levels by diagnosis
g1 <- ggplot(thyroid_data, aes(x = Diagnosis, y = TSH_Level, fill = Diagnosis)) + geom_boxplot() + ggtitle("TSH Levels by Diagnosis")
g2 <- ggplot(thyroid_data, aes(x = Diagnosis, y = T3_Level, fill = Diagnosis)) + geom_boxplot() + ggtitle("T3 Levels by Diagnosis")
g3 <- ggplot(thyroid_data, aes(x = Diagnosis, y = T4_Level, fill = Diagnosis)) + geom_boxplot() + ggtitle("T4 Levels by Diagnosis")
g4 <- ggplot(thyroid_data, aes(x = Diagnosis, y = Nodule_Size, fill = Diagnosis)) + geom_boxplot() + ggtitle("Nodule Size by Diagnosis")
g5 <- ggplot(thyroid_data, aes(x = Diagnosis, y = Age, fill = Diagnosis)) + geom_boxplot() + ggtitle(" Age by Diagnosis")

g1
g2
g3
g4
g5

```


checking why the boxplots are identical, is there variation in their values:
```{r}
thyroid_data %>%
  group_by(Diagnosis) %>%
  summarise(across(c(TSH_Level, T3_Level, T4_Level, Nodule_Size), list(mean = mean, sd = sd, min = min, max = max)))
```


t test to check if theres any statistical significance
```{r}
t.test(TSH_Level ~ Diagnosis, data = thyroid_data)
t.test(T3_Level ~ Diagnosis, data = thyroid_data)
t.test(T4_Level ~ Diagnosis, data = thyroid_data)
t.test(Nodule_Size ~ Diagnosis, data = thyroid_data)
```


# now trying to use Thyroid_Cancer_Risk as the outcome variable instead, testing if theres any significance here
```{r}
# Boxplots for numerical variables by Thyroid_Cancer_Risk
g1 <- ggplot(thyroid_data, aes(x = Thyroid_Cancer_Risk, y = TSH_Level, fill = Thyroid_Cancer_Risk)) + 
  geom_boxplot(alpha = 0.5) + 
  ggtitle("TSH Levels by Thyroid Cancer Risk") +
  theme_minimal()

g2 <- ggplot(thyroid_data, aes(x = Thyroid_Cancer_Risk, y = T3_Level, fill = Thyroid_Cancer_Risk)) + 
  geom_boxplot(alpha = 0.5) + 
  ggtitle("T3 Levels by Thyroid Cancer Risk") +
  theme_minimal()

g3 <- ggplot(thyroid_data, aes(x = Thyroid_Cancer_Risk, y = T4_Level, fill = Thyroid_Cancer_Risk)) + 
  geom_boxplot(alpha = 0.5) + 
  ggtitle("T4 Levels by Thyroid Cancer Risk") +
  theme_minimal()

g4 <- ggplot(thyroid_data, aes(x = Thyroid_Cancer_Risk, y = Age, fill = Thyroid_Cancer_Risk)) + 
  geom_boxplot(alpha = 0.5) + 
  ggtitle("Age by Thyroid Cancer Risk") +
  theme_minimal()

g5 <- ggplot(thyroid_data, aes(x = Thyroid_Cancer_Risk, y = Nodule_Size, fill = Thyroid_Cancer_Risk)) + 
  geom_boxplot(alpha = 0.5) + 
  ggtitle("Nodule Size by Thyroid Cancer Risk") +
  theme_minimal()

g1
g2
g3
g4
g5

```



```{r}
str(thyroid_data)
summary(thyroid_data[, numerical_var])
```




Check correlation between numerical predictors
```{r}
num_df <- thyroid_data %>% select_if(is.numeric)
corr_matrix <- cor(num_df, use = "complete.obs")
corrplot(corr_matrix, method = "color", tl.cex = 0.7)

library(corrr)
library(ggplot2)
library(GGally)



# Convert to numeric if necessary
thyroid_data[numerical_var] <- lapply(thyroid_data[numerical_var], as.numeric)

# Compute correlation matrix
cor_matrix <- cor(thyroid_data[numerical_var], use = "complete.obs", method = "pearson")

# Print correlation matrix
print(cor_matrix)

# Visualize correlation matrix
ggcorrplot::ggcorrplot(cor_matrix, lab = TRUE, title = "Correlation Matrix of Continuous Variables")

```


check multicollinearity (VIF)
```{r}

library(car)
# Run logistic regression instead of linear regression
vif_model <- glm(Diagnosis ~ Family_History + Radiation_Exposure + Iodine_Deficiency + 
                  Smoking + Obesity + Diabetes + Gender + Country + Ethnicity, 
                  data = thyroid_data, family = binomial)

# Calculate VIF
vif_values <- vif(vif_model)
print(vif_values)

```


Now scale the data between 0-1
```{r}
# Scaling (0 to 1)
thyroid_data[numerical_var] <- as.data.frame(lapply(thyroid_data[numerical_var], function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))

summary(thyroid_data[numerical_var])

```

```{r}
# Standardise (Z score normalisation)
thyroid_data[numerical_var] <- as.data.frame(lapply(thyroid_data[numerical_var], function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}))

summary(thyroid_data[numerical_var])
```



```{r}
# GLM on full data
glm_model <- glm(Diagnosis ~ ., data = thyroid_data, family = binomial)

# Display model summary
summary(glm_model)
```

