Predicting Colorectal Cancer

Libraries
```{r}
library(ggplot2) # for all the graphs/ plots
library(tidyverse)
```

Load the dataset
```{r}
data <- read.csv("colorectal_cancer_dataset.csv", stringsAsFactors = TRUE)
summary(data)
str(data)
colnames(data)
```

Categorical variables as factors
```{r}
categorical_vars <- c("Country", "Gender", "Cancer_Stage", "Family_History", "Smoking_History", "Alcohol_Consumption",
                      "Obesity_BMI", "Diet_Risk", "Physical_Activity", "Diabetes", "Inflammatory_Bowel_Disease", "Genetic_Mutation",
                      "Screening_History", "Early_Detection", "Treatment_Type", "Survival_5_years", "Mortality", "Urban_or_Rural",
                      "Economic_Classification", "Healthcare_Access", "Insurance_Status", "Survival_Prediction")
data <- data %>%
  mutate(across(all_of(categorical_vars), as.factor))
```

Continuous variables as numeric
```{r}
numeric_vars <- c("Age", "Tumor_Size_mm","Healthcare_Costs", "Incidence_Rate_per_100K", "Mortality_Rate_per_100K")
data <- data %>%
  mutate(across(all_of(numeric_vars), as.numeric))
```

Check for missing values, there are no missing values.
```{r}
colSums(is.na(data))
```

Create barcharts for all the categorical variables
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# Loop through all categorical variables and create bar plots
for (var in categorical_vars) {
  p <- ggplot(data, aes_string(x = var, fill = "Survival_Prediction")) +
    geom_bar(position = "fill") +  # "fill" shows proportions within each category
    theme_minimal() +
    ggtitle(paste("Impact of", var, "on Survival_Prediction")) +
    ylab("Proportion") +
    xlab(var) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for better readability
  
  print(p)
}
```

Create boxplots for all the numerical variables
```{r}
# Distribution of Survival_Prediction
ggplot(data, aes(x = Survival_Prediction, fill = Survival_Prediction)) + 
  geom_bar() +
  theme_minimal() +
  ggtitle("Survival_Prediction Distribution")

# Plot for each numeric variable
for (var in numeric_vars) {
  
  # Boxplot comparing benign vs malignant
  boxplots <- ggplot(data, aes_string(x = "Survival_Prediction", y = var, fill = "Survival_Prediction")) +
    geom_boxplot() +
    theme_minimal() +
    ggtitle(paste(var, "by Survival_Prediction"))
  print(boxplots)
}
```



Set Benign and Malignant outcomes as 0,1.
There are 357 benign cases and 212 malignant cases in the data set.
```{r}
levels(data$Survival_Prediction)[levels(data$Survival_Prediction) %in% c("No")] <- "0"
levels(data$Survival_Prediction)[levels(data$Survival_Prediction) %in% c("Yes")] <- "1"
summary(data$Survival_Prediction)
```




Check correlation between numerical predictors
```{r}
# Compute the correlation matrix for numeric variables
cor_matrix <- cor(data[, numeric_vars])
print(cor_matrix)


```



```{r}
#Standardise numeric variables using scale() in base R
data[ , numeric_vars] <- scale(data[ , numeric_vars])
library(dplyr)

data <- data %>%
  mutate(across(all_of(numeric_vars), ~ as.numeric(scale(.))))
```



```{r}
# GLM on full data
glm_model <- glm(Survival_Prediction ~ ., data = data, family = binomial)

# Display model summary
summary(glm_model)
```


Check multicolinearity:
```{r}
library(car)
vif_values <- vif(glm_model)
print(vif_values)

```




