
Load the dataset, pick the relevant variables and export as CSV file
Do not run this, run chunk 2 onwards
```{r}
# Load libraries
library(haven)
library(tidyverse)
library(dplyr)

# We got the dataset from https://webfs.oecd.org/pisa2018/index.html

student_file <- "~/Documents/1. uni/Year 3/AT - ST310/group/CY07_MSU_STU_QQQ.sav"

df_student <- read_sav(student_file)

# Step 1: Select Relevant Variables for the project, which will be good for interpretabilitiy
df_student_selected <- df_student %>%
  dplyr::select(CNTSTUID,    # Student ID
         PV1MATH,     # Math Score
         PV1READ,     # Reading Score
         PV1SCIE,     # Science Score
         ESCS,        # Socioeconomic Status
         ST004D01T,   # Gender
         ST005Q01TA,  # Highest level of schooling completed by your mother
         ST007Q01TA,  # Highest level of schooling completed by your father?
         ST011Q04TA,  # In your home: A computer you can use for school work
         ST011Q06TA,  # In your home: A link to the Internet
         ST012Q09NA,  # How many in your home: Musical instruments (e.g. guitar, piano)
         IC006Q01TA,  # During a typical weekday, how long do you use Internet outside school
         WB150Q01HA,  # How is your health?
         PA042Q01TA,  # Annual household income?
         EFFORT1,     # How much effort did you put into this test?
         IMMIG,       # Index Immigration status
         ST034Q06TA,  # Thinking about your school: I feel lonely at school.
         ST176Q03IA,  # How often involved in: Reading online news
         ST175Q01IA,  # About how much time do you usually spend reading for enjoyment?
         ST022Q01TA,  # What language do you speak at home most of the time?
         ST012Q02TA,  # How many in your home: Cars
         ST016Q01NA,  # Overall, how satisfied are you with your life as a whole these days?
         AGE,         # Age
         TMINS,       # Learning time (minutes per week) - in total
         SCCHANGE,    # Number of school changes
         COMPETE,     # Competitiveness (WLE)
         GFOFAIL,     # General fear of failure (WLE)
         STUBMI       # BMI of student
       
  )

# Step 2: Handle Missing Data
# Drop columns which has over 30% of missing values
df_student_cleaned <- df_student_selected %>%
  dplyr::select(where(~ mean(is.na(.)) < 0.3))

# Fill small missing values with median for numerical columns
df_student_cleaned <- df_student_cleaned %>%
  mutate(across(where(is.numeric), ~ replace_na(., median(., na.rm = TRUE))))

# Convert categorical variables to factors
df_student_cleaned <- df_student_cleaned %>%
  mutate(across(where(is.character), as.factor))

# Step 4: Save the Cleaned and Merged Dataset as CSV onto my desktop
output_file <- "~/Desktop/Final_310_dataset.csv"
write.csv(df_student_cleaned, output_file, row.names = FALSE)

# View Summary of Merged Data
summary(df_student_cleaned)
```


Load the data file
```{r}
data <- read.csv("Final_310_dataset.csv", stringsAsFactors = TRUE)
summary(data)
str(data)
colnames(data)
```

Categorical variables as factors
```{r}
categorical_vars <- c("ST004D01T", "ST005Q01TA", "ST007Q01TA", "ST011Q04TA",
                      "ST011Q06TA", "ST012Q09NA", "EFFORT1", "IMMIG", "ST034Q06TA",
                      "ST176Q03IA", "ST175Q01IA", "ST022Q01TA", "ST012Q02TA", "ST016Q01NA" 
                      )
data <- data %>%
  mutate(across(all_of(categorical_vars), as.factor))
```

Continuous variables as numeric
```{r}
numeric_vars <- c("PV1MATH","PV1READ","PV1SCIE", "ESCS", "AGE", "COMPETE", "GFOFAIL")
data <- data %>%
  mutate(across(all_of(numeric_vars), as.numeric))
```

Check for missing values, there are no missing values.
```{r}
colSums(is.na(data))
```



Create boxplots for all the categorical variables against maths score (PV1MATH)
```{r}
for (var in categorical_vars) {
    print(ggplot(data, aes(x = get(var), y = PV1MATH, fill = get(var))) +
        geom_boxplot() +
        labs(title = paste("PV1MATH by", var),
             x = var,
             y = "PV1MATH") +
        theme_minimal()) 
}
```



Create scatterplots for all the continuous variables against maths score (PV1MATH)
```{r}
scatter_plots <- lapply(numeric_vars, function(var) {
  ggplot(data, aes_string(x = var, y = "PV1MATH")) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(title = paste("PV1MATH vs", var),
         x = var,
         y = "PV1MATH") +
    theme_minimal()
})

scatter_plots

```





Check correlation between numerical predictors
```{r}
# Compute the correlation matrix for numeric variables
cor_matrix <- cor(data[, numeric_vars])
print(cor_matrix)


```




Run regression
```{r}
# LM on full data
lm_model <- lm(PV1MATH ~ ., data = data, family = binomial)

# Display model summary
summary(lm_model)
display(lm_model)
Anova(lm_model)
```




Check multicolinearity:
```{r}
library(car)
vif_values <- vif(lm_model)
print(vif_values)

```



